---
title: "Oyster ploidy field study survival analysis"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
  theme: sky
  code_folding: hide
---

Survival analysis to the end of 2023 for oyster ploidy project



## Read in the data and create basic tables and graphs

Load the required packages. Using survival package that was updated in August 2023

```{r, results=FALSE, warning=FALSE, error=FALSE, message=FALSE, }
library(tidyverse)
library(here)
library(survival)
library(ggfortify)
library(ggsurvfit)
library(survminer)
library(gridExtra)
library(frailtyEM)
library(coxme)
```

Read in the mortality data from the field experiments

```{r, warning=FALSE, error=FALSE, message=FALSE}
survdat <- read.csv(here("Data", "MAY_2023_JULY_2024_MORTSBYBAG.csv"))
```

## Data Wrangling

Correct the data structure

```{r, warning=FALSE, error=FALSE, message=FALSE}
survdat <- survdat %>% mutate_at(c("Site", "Bag_Col", "Month", "Bag_no", "Month_no", "Year", "Month_year"), as.factor)
str(survdat)
```

Create a new data sheet that calculates the number of survivors at each sampling event (as opposed to the number of moralities)

```{r, warning=FALSE, error=FALSE, message=FALSE, echo=FALSE}
widebagmort <- pivot_wider(survdat[-c(6, 3, 7, 9)], names_from = Month_year, values_from = Mort_count)#C6 is removed because this is numerical month
#replace NA in the mortality columns with 0
widebagmort <- widebagmort %>%
  mutate_at(vars(May_2023, June_2023, July_2023, August_2023, September_2023, October_2023, May_2024, June_2024, July_2024), ~replace_na(., 0))
widebagmort <- widebagmort %>% 
  group_by(Site, Bag_Col, Bag_no, )
summarise(widebagmort)
widebagmort$April_2023_survivors <- 50
widebagmort$May_2023_survivors <- widebagmort$April_2023_survivors-widebagmort$May_2023
widebagmort$June_2023_survivors <- widebagmort$May_2023_survivors-widebagmort$June_2023
widebagmort$July_2023_survivors <- widebagmort$June_2023_survivors-widebagmort$July_2023
widebagmort$August_2023_survivors <- widebagmort$July_2023_survivors-widebagmort$August_2023
widebagmort$September_2023_survivors <- widebagmort$August_2023_survivors-widebagmort$September_2023
widebagmort$October_2023_survivors <- widebagmort$September_2023_survivors-widebagmort$October_2023
widebagmort$May_2024_survivors <- widebagmort$October_2023_survivors-widebagmort$May_2024
widebagmort$June_2024_survivors <- widebagmort$May_2024_survivors-widebagmort$June_2024
widebagmort$July_2024_survivors <- widebagmort$June_2024_survivors-widebagmort$July_2024

#Pivit back to longer
no_survivors <- widebagmort %>% pivot_longer(cols=c(13:22),#######################As months are added make sure their columns are selected - change to an auto thing*************
                                                         names_to='Month_survivor',
                                                         values_to='No_survivors') %>% 
  select("Site", "Bag_Col", "Bag_no",'Month_survivor','No_survivors')
no_survivors$Month_survivor <- as.factor(no_survivors$Month_survivor)
no_survivors
```

Double check that all bags are accounted for. There should be 8 bags per colour site per sampling event = 56 as of October 2023

```{r, echo=FALSE}
no_survivors %>% group_by(Site, Bag_Col) %>% summarise(nbags=n())
#add the date column on here

no_survivors <- no_survivors %>%
  mutate(Month_survivor = as.character(Month_survivor),
         Month = sub("_.*", "", Month_survivor),  # Extract month
         Year = sub(".*_(\\d+)_survivors", "\\1", Month_survivor),  # Extract year
         Month_no = match(Month, month.name),  # Convert month name to month number
         Date = ymd(paste0(Year, "-", sprintf("%02d", Month_no), "-01")))  # Create date
```

Summarise the number of survivors per site per sampling event

```{r, echo=FALSE}
mean_survival_proportion <- no_survivors %>% 
  group_by(Site, Bag_Col, Date) %>% 
  summarise(meansurv=mean(No_survivors), semsurv=sd(No_survivors)/sqrt(n()))

mean_survival_proportion
```

Create a basic survival plot taht shows the number of survivors at each sampling event

```{r, echo=FALSE}
mean_survival_proportion %>% ggplot(aes(x=Date, y=meansurv, colour = Bag_Col, group=Bag_Col))+geom_line()+
  geom_errorbar(aes(ymin=meansurv-semsurv, ymax=meansurv+semsurv))+
  geom_line()+facet_grid(~Site)+
  xlab("Month")+ylab("Proportional Survival")+
  theme_bw()+ggtitle("Number of survivors since experiment start")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Need to turn the data into individual survival data. Each oyster should be listed with either alive or dead at each sampling point. Each oyster will need an ID number. 8 Cages per ploidy per site were monitored for mortality so a total of 400 oysters per ploidy per site at the start. Total of 4800 oysters. 

This is a very clunky way to do this and needs to be streamlined - a for loop might work?

```{r, echo = FALSE, warning=FALSE, error=FALSE, message=FALSE}
#the uncount function will work here
#Count just the April individuals to get ID numbers for each individual at the start
no_survivorsexpanded_April <- no_survivors %>% filter(Month_survivor == 'April_2023_survivors') %>% 
  uncount(No_survivors) #repeats each row by the number of times in April_survivors col
#Give each individual an ID number from 1-50
no_survivorsexpanded_April$OYS_ID <- rep(1:50, times = nrow(no_survivorsexpanded_April)/50) #Gives each a number
no_survivorsexpanded_April$OYS_ID <- paste0(no_survivorsexpanded_April$Site, no_survivorsexpanded_April$Bag_Col,
                                            no_survivorsexpanded_April$Bag_no, "_", no_survivorsexpanded_April$OYS_ID)
no_survivorsexpanded_April$April_status <- 0 #gives everyone who is alive a status of 0
##Do this for each month ##MAY##
no_survivorsexpanded_May <- no_survivors %>% filter(Month_survivor == 'May_2023_survivors') %>% 
  uncount(No_survivors) #repeats each row by the number of times in May_survivors col
no_survivorsexpanded_May <- no_survivorsexpanded_May %>% #Gives each individual in each bag a number (max of 50)
  group_by(Site, Bag_Col, Bag_no) %>% 
  mutate(OYS_ID = (row_number() - 1) %% 50 + 1)
no_survivorsexpanded_May$OYS_ID <- paste0(no_survivorsexpanded_May$Site, no_survivorsexpanded_May$Bag_Col,
                                          no_survivorsexpanded_May$Bag_no, "_", no_survivorsexpanded_May$OYS_ID)
no_survivorsexpanded_May$May_status <- 0 #gives everyone who is alive a status of 0
ind_surv_dat <- left_join(select(no_survivorsexpanded_April, Site, Bag_Col, Bag_no, OYS_ID, April_status),
                          select(no_survivorsexpanded_May, OYS_ID, May_status),
                          by = "OYS_ID")#Join this to the april data

##Do this for each month ##June##
no_survivorsexpanded_June <- no_survivors %>% filter(Month_survivor == 'June_2023_survivors') %>% 
  uncount(No_survivors) #repeats each row by the number of times in May_survivors col
no_survivorsexpanded_June <- no_survivorsexpanded_June %>% #Gives each individual in each bag a number (max of 50)
  group_by(Site, Bag_Col, Bag_no) %>% 
  mutate(OYS_ID = (row_number() - 1) %% 50 + 1)
no_survivorsexpanded_June$OYS_ID <- paste0(no_survivorsexpanded_June$Site, no_survivorsexpanded_June$Bag_Col,
                                          no_survivorsexpanded_June$Bag_no, "_", no_survivorsexpanded_June$OYS_ID)
no_survivorsexpanded_June$June_status <- 0 #gives everyone who is alive a status of 0
ind_surv_dat <- left_join(ind_surv_dat,
                          no_survivorsexpanded_June,
                          by = "OYS_ID") #Join this to the existing data

##Do this for each month ##July##
no_survivorsexpanded_July <- no_survivors %>% filter(Month_survivor == 'July_2023_survivors') %>% 
  uncount(No_survivors) #repeats each row by the number of times in May_survivors col
no_survivorsexpanded_July <- no_survivorsexpanded_July %>% #Gives each individual in each bag a number (max of 50)
  group_by(Site, Bag_Col, Bag_no) %>% 
  mutate(OYS_ID = (row_number() - 1) %% 50 + 1)
no_survivorsexpanded_July$OYS_ID <- paste0(no_survivorsexpanded_July$Site, no_survivorsexpanded_July$Bag_Col,
                                          no_survivorsexpanded_July$Bag_no, "_", no_survivorsexpanded_July$OYS_ID)
no_survivorsexpanded_July$July_status <- 0 #gives everyone who is alive a status of 0
ind_surv_dat <- left_join(ind_surv_dat,
                          no_survivorsexpanded_July, 
                          by = "OYS_ID") #Join this to the existing data

##Do this for each month ##August##
no_survivorsexpanded_August <- no_survivors %>% filter(Month_survivor == 'August_2023_survivors') %>% 
  uncount(No_survivors) #repeats each row by the number of times in May_survivors col
no_survivorsexpanded_August <- no_survivorsexpanded_August %>% #Gives each individual in each bag a number (max of 50)
  group_by(Site, Bag_Col, Bag_no) %>% 
  mutate(OYS_ID = (row_number() - 1) %% 50 + 1)
no_survivorsexpanded_August$OYS_ID <- paste0(no_survivorsexpanded_August$Site, no_survivorsexpanded_August$Bag_Col,
                                          no_survivorsexpanded_August$Bag_no, "_", no_survivorsexpanded_August$OYS_ID)
no_survivorsexpanded_August$August_status <- 0 #gives everyone who is alive a status of 0
ind_surv_dat <- left_join(ind_surv_dat,
                          no_survivorsexpanded_August, 
                          by = "OYS_ID") #Join this to the existing data

##Do this for each month ##September##
no_survivorsexpanded_September <- no_survivors %>% filter(Month_survivor == 'September_2023_survivors') %>% 
  uncount(No_survivors) #repeats each row by the number of times in May_survivors col
no_survivorsexpanded_September <- no_survivorsexpanded_September %>% #Gives each individual in each bag a number (max of 50)
  group_by(Site, Bag_Col, Bag_no) %>% 
  mutate(OYS_ID = (row_number() - 1) %% 50 + 1)
no_survivorsexpanded_September$OYS_ID <- paste0(no_survivorsexpanded_September$Site, no_survivorsexpanded_September$Bag_Col,
                                          no_survivorsexpanded_September$Bag_no, "_", no_survivorsexpanded_September$OYS_ID)
no_survivorsexpanded_September$September_status <- 0 #gives everyone who is alive a status of 0
ind_surv_dat <- left_join(ind_surv_dat,
                          no_survivorsexpanded_September, 
                          by = "OYS_ID") #Join this to the existing data

##Do this for each month ##October##
no_survivorsexpanded_October <- no_survivors %>% filter(Month_survivor == 'October_2023_survivors') %>% 
  uncount(No_survivors) #repeats each row by the number of times in May_survivors col
no_survivorsexpanded_October <- no_survivorsexpanded_October %>% #Gives each individual in each bag a number (max of 50)
  group_by(Site, Bag_Col, Bag_no) %>% 
  mutate(OYS_ID = (row_number() - 1) %% 50 + 1)
no_survivorsexpanded_October$OYS_ID <- paste0(no_survivorsexpanded_October$Site, no_survivorsexpanded_October$Bag_Col,
                                          no_survivorsexpanded_October$Bag_no, "_", no_survivorsexpanded_October$OYS_ID)
no_survivorsexpanded_October$October_status <- 0 #gives everyone who is alive a status of 0
ind_surv_dat <- left_join(ind_surv_dat,
                          no_survivorsexpanded_October, 
                          by = "OYS_ID") #Join this to the existing data

##Do this for each month ##May 2024##
no_survivorsexpanded_May_24 <- no_survivors %>% filter(Month_survivor == 'May_2024_survivors') %>% 
  uncount(No_survivors) #repeats each row by the number of times in May_survivors col
no_survivorsexpanded_May_24 <- no_survivorsexpanded_May_24 %>% #Gives each individual in each bag a number (max of 50)
  group_by(Site, Bag_Col, Bag_no) %>% 
  mutate(OYS_ID = (row_number() - 1) %% 50 + 1)
no_survivorsexpanded_May_24$OYS_ID <- paste0(no_survivorsexpanded_May_24$Site, no_survivorsexpanded_May_24$Bag_Col,
                                          no_survivorsexpanded_May_24$Bag_no, "_", no_survivorsexpanded_May_24$OYS_ID)
no_survivorsexpanded_May_24$May_24_status <- 0 #gives everyone who is alive a status of 0
ind_surv_dat <- left_join(ind_surv_dat,
                          no_survivorsexpanded_May_24, 
                          by = "OYS_ID") #Join this to the existing data
##Do this for each month ##June 2024##
no_survivorsexpanded_June_24 <- no_survivors %>% filter(Month_survivor == 'June_2024_survivors') %>% 
  uncount(No_survivors) #repeats each row by the number of times in June_survivors col
no_survivorsexpanded_June_24 <- no_survivorsexpanded_June_24 %>% #Gives each individual in each bag a number (max of 50)
  group_by(Site, Bag_Col, Bag_no) %>% 
  mutate(OYS_ID = (row_number() - 1) %% 50 + 1)
no_survivorsexpanded_June_24$OYS_ID <- paste0(no_survivorsexpanded_June_24$Site, no_survivorsexpanded_June_24$Bag_Col,
                                          no_survivorsexpanded_June_24$Bag_no, "_", no_survivorsexpanded_June_24$OYS_ID)
no_survivorsexpanded_June_24$June_24_status <- 0 #gives everyone who is alive a status of 0
ind_surv_dat <- left_join(ind_surv_dat,
                          no_survivorsexpanded_June_24, 
                          by = "OYS_ID") #Join this to the existing data
##Do this for each month ##JULY 2024##
no_survivorsexpanded_July_24 <- no_survivors %>% filter(Month_survivor == 'July_2024_survivors') %>% 
  uncount(No_survivors) #repeats each row by the number of times in July_survivors col
no_survivorsexpanded_July_24 <- no_survivorsexpanded_July_24 %>% #Gives each individual in each bag a number (max of 50)
  group_by(Site, Bag_Col, Bag_no) %>% 
  mutate(OYS_ID = (row_number() - 1) %% 50 + 1)
no_survivorsexpanded_July_24$OYS_ID <- paste0(no_survivorsexpanded_July_24$Site, no_survivorsexpanded_July_24$Bag_Col,
                                          no_survivorsexpanded_July_24$Bag_no, "_", no_survivorsexpanded_July_24$OYS_ID)
no_survivorsexpanded_July_24$July_24_status <- 0 #gives everyone who is alive a status of 0
ind_surv_dat <- left_join(ind_surv_dat,
                          no_survivorsexpanded_July_24, 
                          by = "OYS_ID") #Join this to the existing data



#Now just select the important columns                      
ind_surv_dat <- ind_surv_dat %>% select(Site.x, Bag_Col.x, Bag_no.x, OYS_ID,
                                        April_status,
                                        May_status,
                                        June_status,
                                        July_status,
                                        August_status,
                                        September_status,
                                        October_status,
                                        May_24_status,
                                        June_24_status,
                                        July_24_status)
ind_surv_dat
```

Pivot this data longer to get the correct formayt for survival analyses

```{r, echo=FALSE}
#Rename 2023 columns by month_year
ind_surv_dat <- ind_surv_dat %>% rename(
  April_23_status = April_status,
  May_23_status = May_status,
  June_23_status = June_status,
  July_23_status = July_status,
  August_23_status = August_status,
  September_23_status = September_status,
  October_23_status = October_status
)

survival_dat <- ind_surv_dat %>% pivot_longer(cols=c(April_23_status, May_23_status, June_23_status,
                                              July_23_status, August_23_status, September_23_status, October_23_status, May_24_status, June_24_status, July_24_status),
                                              names_to = "Month",
                                              values_to = "event")
survival_dat <- survival_dat %>%  mutate(Month = str_replace(Month, "_status", ""))#remove the status from the data holes (cells)
#Add a numeric month number
survival_dat$Date <- parse_date_time(survival_dat$Month, orders = "my", tz = "UTC")
start_date <- ymd("2023-04-01")# Define the starting month
survival_dat$month_no <- interval(start_date, survival_dat$Date) %/% months(1) + 4

#replace the NAs in the event column with 1
survival_dat <- survival_dat %>% 
  mutate(event = ifelse(is.na(event), 1, event))

#I only want to keep the first instance of a 1 for each individual.
survival_dat <- survival_dat %>%
  group_by(OYS_ID) %>%
  mutate(event = replace(event, cumsum(event == 1) > 1, NA)) %>% 
  mutate(event = ifelse(event == 0 | cumsum(event == 1) == 1, event, 0))

```

## Survival analyses

### Survial plots for pooled data (no random effects of bag number)

Created Kaplan-Meyer survival plots for all data pooled across sites and ploidys. They do not consider the potential random effects of Bag within sites. 

Create a survival plot faceted by Site

```{r, echo = FALSE}
pooled_site_survplot <- survival_dat %>% 
 ggsurvplot_facet(survfit(Surv(month_no, event, type = "right") ~ Bag_Col.x , data = .),
            facet.by = "Site.x",
            data = .,
            risk.table = FALSE, pval = FALSE, conf.int = TRUE, 
            xlim = c(4, 19), ylim = c(0.65, 1),
            legend.labs = c("Diploid", "Induced Triploid", "Mated Triploid"),
            xlab = "Month",
            legend.strata = FALSE, 
            legend.title = "Ploidy")+
  scale_x_continuous(breaks = seq(4, 19, by = 1))+
  theme_bw(base_size = 15)+
  labs(legend = "ploidy")+
  scale_color_manual(values = c("blue", "green", "red"))+
  scale_fill_manual(values = c("blue", "green", "red"))
pooled_site_survplot

```

Create a survival plot faceted by ploidy

```{r}
pooled_ploidy_survplot <- survival_dat %>% 
 ggsurvplot_facet(survfit(Surv(month_no, event, type = "right") ~ Site.x , data = .),
            facet.by = "Bag_Col.x",
            data = .,
            risk.table = FALSE, pval = FALSE, conf.int = TRUE,
            xlab = ("Month"),
            xlim = c(4, 19), ylim = c(0.6, 1))
pooled_ploidy_survplot
```

These plots show that there appeared to be an effect of site on survival throughout summer of 2023

Now I am going to create a cox proportional hazards model.


```{r, echo=FALSE}
surv_field <- with(survival_dat, Surv(month_no, event, type = "right"))#For some reason piping does not work here

coxph_pooled <-  coxph(surv_field ~ Site.x * Bag_Col.x, data = survival_dat)
anova(coxph_pooled)

```

###Coxme package to examine differences
Using the coxme oackage allows us to introcuce a random effect of cage. This may not be essential, but at least we can check

```{r}
coxph_randomeff <- coxme(surv_field ~ Site.x*Bag_Col.x + (1|Bag_no.x), data = survival_dat)
```
We comapeed the two models (one with random effect of bag an one without)

```{r}
anova(coxph_pooled, coxph_randomeff)
```

The model with the random effect of bag is better than the one without this random effect, so we will use the model with teh random effect.

```{r}
pairwise_coxph_randomeff <- emmeans(coxph_randomeff, ~ Site.x*Bag_Col.x)
pairs_ploidy <- pairs(pairwise_coxph_randomeff, simple = "Site.x", adjust = "Tukey")
pairs_CO2 <- pairs(pairwise_coxph_randomeff, simple = "ag_Col.x", adjust = "Tukey")
```





There was a significant ploidy*site interaction in this analysis.
Significant differences existed between each of the sites and 


### Include the random effects of cage

Following Paul's example from the crab pH analyses I am going to use the frailtyEM package (@balan2019) to do a mixed effects analysis. I am including cage as a random effect because it acts as a grouping variable and we really don't care that much about the impacts of cage in our interpretation. 

  






Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
