---
title: "Oyster ploidy field study survival analysis"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
  theme: sky
  code_folding: hide
---

Load the required packages

```{r}
library(tidyverse)
library(here)
library(survival)
```

Read in the mortality data from the field experiments

```{r}
survdat <- read.csv(here("Data", "MAY_SEPT_2023_MORTSBYBAG.csv"))
```

Ensure data structure is correct

```{r}
survdat <- survdat %>% mutate_at(c("Site", "Bag_Col", "Month", "Bag_no", "Month_no"), as.factor)
```

Inspect the data structure

```{r}
str(survdat)
```

Create a new data sheet that calculates teh number of survivors at each sampling event (as opposed to the number of mortalities)

```{r}
widebagmort <- pivot_wider(survdat[-c(6)], names_from = Month, values_from = Mort_count)
#replace NA in the mortality columns with 0
widebagmort <- widebagmort %>%
  mutate_at(vars(May, June, July, Aug, September), ~replace_na(., 0))
widebagmort$May_survivors <- 50-widebagmort$May
widebagmort$June_survivors <- widebagmort$May_survivors-widebagmort$June
widebagmort$July_survivors <- widebagmort$June_survivors-widebagmort$July
widebagmort$Aug_survivors <- widebagmort$July_survivors-widebagmort$Aug
widebagmort$Sept_survivors <- widebagmort$Aug_survivors-widebagmort$September

no_survivors <- widebagmort %>% pivot_longer(cols=c(9:12),
                                                         names_to='Month_survivor',
                                                         values_to='No_survivors') %>% 
  select("Site", "Bag_Col", "Bag_no",'Month_survivor','No_survivors')

no_survivors
```

Double check that all bags are accounted for. There should be 8 bags per site

```{r}
widebagmort %>% group_by(Site, Bag_Col) %>% summarise(nbags=n())
```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
